---
title: "Star Wars musical ecology"
author: "Chris Hamm"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    toc_depth: 3
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center", cache = TRUE)
```

## Introduction

### Star Wars Oxygen
  I love [*Star Wars.*](http://www.starwars.com/) I love the story telling and fantasy, but I especially love the music. John Williams is amazing. There was a podcast out there called [*Star Wars Oxygen*](http://www.rebelforceradio.com/star-wars-oxygen/) that covered the music of *Star Wars* and it was one of my favorite podcasts of all time. [Jimmy Mac](http://www.rebelforceradio.com/jimmy-mac/) hosted while voice actor, musicician, and composer [David W. Collins](http://starwars.wikia.com/wiki/David_W._Collins) broke down the scores for the films we know and love in a way that gave me a new appreciation for the films. I say there *was a podcast* because the podcast went dark following the release of Rogue One. After 38 wonderful volumes the podcast just wasn't updated any more and we the fans have not heard anything about why they stopped producing the show. 

### Species diversity
  I also love statistics and ecology, which is the study of how organisms relate to each other and their environments. One exciting area of research deals with **diversity.** We can use statistics to figure out how many things live in a certain area and compare how different habitats are similar or different to one another. In order to conduct an analysis like this you need a "count matrix," which has **habitats** on rows and **species** in columns. The cells are filled in with counts of how many of each **species** is found in each habitat. An example of a count matrix could look like this:
  
  ||*Danaus plexippus*|*Vanessa cardui*|*Adelpha bredowii*|
  |:--:|:--:|:--:|:--:|
  |Donner Pass|5|6|0|
  |Sierraville|4|2|2|
  |Davis|0|0|3|
Table: Example of a count matrix where each row represents a habitat and each column represents a species. The cells are filled in with counts of the number of each species observed at each habitat.

In this example, we can see that Donner Pass and Sierraville are similar to each other for two species. Also Davis and Sierraville are somewhat similar to each other because they have one species in common. If we were going to group these sites based on similarity, Donner Pass and Sierraville would be more similar to each other than to Davis.

```{r preliminaries, include = FALSE}
set.seed(896235)

library("vegan")
library("vegetarian")
library("spaceMovie")
```
```{r Toy_example, include = FALSE}
Toy_data <- data.frame(D_plex = c(5, 4, 0), V_cardui = c(6, 2, 2), A_bredowii =  c(0, 0, 3), row.names = c("Donner_Pass", "Sierraville", "Davis"))

Toy_dist <- vegdist(Toy_data, method = "jaccard", diag = FALSE)
Toy_clust <- hclust(Toy_dist, "ward.D")
```

If we plot these relationships as a tree (after some statistical wankery) we see that Donner Pass and Sierraville appear close together with Davis far apart from them. 

```{r Toy_clust, echo = FALSE, fig.cap = "Cluster plot of the toy example referred to above."}
plot(Toy_clust, main = "", xlab = "", sub = "")
```

Please note here that I have created this page using `RMardown` in `RStudio`. All of the code and data used to create this post are freely available through this project's [`github` repository]().

### Star Wars musical ecology

During the *Star Wars Oxygen* podcast, David W. Collins began what he called his "theme tracker," which was essential a spreadsheet of the number of times a theme played per film. 

David W. Collins made a count matrix. 

We can apply statistics to *Star Wars*!!!! Oh happy day!!!

## The data

To reverse enginerer the theme tracker I listened back through all of the *Star Wars Oxygen* episode with pencil and paper ready. I made note of how often a theme was played during a particular film every time Mr. Collins mentioned it. In some instances, I had to get a bit of help and I read the breakdowns and threads from these sites:

* [jwfan.com](http://www.jwfan.com/)
* [*Star Wars* wikia](http://starwars.wikia.com/wiki/Binary_Sunset)
* [TheForce.net](http://boards.theforce.net/threads/force-theme-archive-just-how-many-tracks-is-the-force-theme-in-anyways.8458343/)

This was especially helpful when going the **AOTC**, which had a lot of music edits.

The data I ended up with, and which is used here had:

* **7** rows - one for each film ("ecosystem")
* **35** columns - one for each theme ("species")

```{r load_data, include = FALSE}
SW_data <- read.csv("../data/DWC_theme_tracker.csv", header = TRUE, row.names = 1)
dim(SW_data)
```

These data could be incomplete and are in need of improvement. I am particularly concerned by the lack of "rare" themes in the data set. Rare things are important in ecology. There are two ways you could contribute:

* Edit a Google Doc
* Submit a PR through `github` (for those with technical skills). 

## Analysis

Now we'll make a tree depicting the relationships between the seven films of the *Star Wars* saga just as we did in the above example. 

```{r SW_clust, include = FALSE}
SW_dist <- vegdist(SW_data, method = "jaccard", diag = FALSE, na.rm = TRUE)

SW_clust <- hclust(SW_dist, method = "ward.D")
```

A prediction on the clustering analysis. The three original trilogy films will cluster together and the pre-quels will cluster together. I think that TFA will be more similar, musically, to the original trilogy than the pre-quels. 

```{r SW_plot, echo = FALSE, fig.cap = "Clustering of the Star Wars films based on the their musical theme counts."}
plot(SW_clust, main = "", xlab = "", sub = "")
```