---
title: "Star Wars musical ecology"
author: "Chris Hamm"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    toc_depth: 3
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center", cache = TRUE)
```

## Introduction

In this post I apply some standard ecological statistics to the music of *Star Wars*. Doing so allows me to combine multiple things that I love. First, I'll walk the reader through the background and the podcast that prompted me to think I could apply statistics to *Star Wars* music. Then I give the reader some background on the type of stats I will apply and give a toy example. Then I go to town and do my best explain the results in plain language, but I do assume that you know what *Star Wars* is and that is has music. If I have failed to describe any of this in plain language please let me know so that I can edit the post and make this more approachable.

A short aside, I have created this page using the program `R` and the `RMarkdown` package in `RStudio`. For those curious, all of the code and data used to create this exact post are freely available through this project's [`github` repository](https://github.com/butterflyology/Star_Wars_musical_ecology).


### Star Wars Oxygen

I **love** [*Star Wars.*](http://www.starwars.com/) I love the story telling and fantasy, but I especially love the music. John Williams is amazing. There was a podcast out there called [*Star Wars Oxygen*](http://www.rebelforceradio.com/star-wars-oxygen/) that covered the music of *Star Wars* and it was one of my favorite podcasts of all time. [Jimmy Mac](http://www.rebelforceradio.com/jimmy-mac/) hosted while voice actor, musician, and composer [David W. Collins](http://starwars.wikia.com/wiki/David_W._Collins) broke down the scores for the films we know and love in a way that gave me a new appreciation for the films. I say there *was a podcast* because the podcast went dark following the release of **Rogue One**. After 38 wonderful volumes the podcast just wasn't updated any more and we the fans have not heard anything about why they stopped producing the show. 


### Species diversity
  I also love statistics and ecology, which is the study of how organisms relate to each other and their environments. One exciting area of research deals with **species diversity**, which is how many species are found in and among sites. We can use statistics to figure out how many things live in a certain area and compare how habitats are similar or different from one another. In order to conduct an analysis like this you need a **count matrix**, which has **habitats** in the rows and **species** in the columns. The cells are filled in with counts of how many of each **species** is found in each **habitat**. An example of a count matrix could look like this, where butterfly **species** are in columns and different **habitats** are in rows:


```{r preliminaries, include = FALSE}
set.seed(896235)

library("vegan")
library("vegetarian")
library("MASS")
library("spaceMovie")
library("tidyverse")
library("plotly")
library("kableExtra")
```


```{r Toy_example, include = FALSE}
Toy_data <- data.frame(Danaus_plexxipus = c(5, 4, 0), Vanessa_cardui = c(6, 2, 2), Adelpha_bredowii =  c(0, 0, 3), row.names = c("Donner_Pass", "Sierraville", "Davis"))

Toy_dist <- vegdist(Toy_data, method = "jaccard", diag = FALSE)
Toy_clust <- hclust(Toy_dist, "ward.D")
```


```{r table_1, }
Genus_species <- c("*Danaus plexippus*", "*Vanessa cardui*", "*Adelpha bredowii*")

knitr::kable(x = kable, format = "html", align = "c", col.names = Genus_species)
```



**Table 1**: Example of a **count matrix** where each row represents a **habitat** and each column represents a **species**. The cells are filled with counts of the number of each species observed at each habitat.
  
  ||*Danaus plexippus*|*Vanessa cardui*|*Adelpha bredowii*|
  |:--:|:--:|:--:|:--:|
  |Donner Pass|5|6|0|
  |Sierraville|4|2|2|
  |Davis|0|0|3|


In this example, we can see that Donner Pass and Sierraville are similar to each other for two species (*D. plexippus* and *V. cardui*). Also Davis and Sierraville are somewhat similar to each other because they have one species in common (*V. cardui* and *A. bredowii*). If we were going to group these habitats based on similarity, Donner Pass and Sierraville would be more similar to each other than to Davis.

If we apply some statistics and then plot these relationships  in the form of a "tree" (**Figure 1**), where similar habitats are connected by a "branch," we see that Donner Pass and Sierraville are most similar to each other (they are connected by a branch). This makes sense, because Donner Pass and Sierraville are about 50 km apart, while both of those sites are ~160 km from Davis.

```{r Toy_clust, echo = FALSE, fig.cap = "**Figure 1**: Cluster plot of the toy example referred to above."}
plot(Toy_clust, main = "", xlab = "", sub = "", lwd = 2)
```


### Star Wars musical ecology

During the *Star Wars Oxygen* podcast, David W. Collins began what he called his "theme tracker," which was essential a spreadsheet of the number of times a theme played per film. Each theme was a column. Each film was a row.

David W. Collins made a **count matrix**.

We can use statistics on a **count matrix**.

We can apply statistics to *Star Wars*!!!! Oh happy day!!!

For the purposes of this exercise, the **theme** will take the place of **species**. What is a theme? A theme, also known as a leitmotif, is a piece of music that is used to represent a character. Remember the music from **A New Hope** that was playing when Luke Skywalker was looking into the binary sunset of Tatooine? The music that played during that scene was called *Obi Wan Kenobi's theme*. In all films since **A New Hope** that music has been called *The Force theme*. The *Imperial March* is a theme; as is the eponymous *Rey's theme* that plays almost every time you see Rey. The fact that a theme plays (almost) every time you see a character is key. So, Yoda has a theme, Kylo Ren has two themes, and Leia has a theme. 


## The data

I was unable to get my very own copy of Mr. Collins theme tracker. No worries. To reverse engineer the theme tracker I listened back through all of the *Star Wars Oxygen* episodes with pencil and paper ready. I made note of how often a theme was played during a particular film every time Mr. Collins mentioned it. In some instances, I had to get a bit of help so I watched the films and made notes of all the times I heard a theme. I also read the breakdowns and threads from these sites:

* [jwfan.com](http://www.jwfan.com/)
* [*Star Wars* wikia](http://starwars.wikia.com/wiki/Binary_Sunset)
* [TheForce.net](http://boards.theforce.net/threads/force-theme-archive-just-how-many-tracks-is-the-force-theme-in-anyways.8458343/)

This was especially helpful when going through **Attack of the Clones**, which had a lot of music edits.

Then I attempted my own impression of David W. Collins and *Star Wars Oxygen* and went through **Rogue One** three times and counted each instance of what I thought was a "theme." At the time of writing this (2017-12-22) **The Last Jedi** has been out for a week and I have seen  four time. The last two viewings I took a piece of paper and a pencil with me to note every time I heard a theme. I have almost certainly missed some things because I am not a trained musician and I could have considered themes to be separate entities when they were they were actually part of the same leitmotifs. Regardless, I stand behind the numbers I present for **Rogue One** and **The Last Jedi**, but please reach out if you find errors.

The data I ended up with, and which are used here, had:

* **9** rows - one for each film ("ecosystem")
* **52** columns - one for each theme ("species")


```{r load_data, include = FALSE}
SW_data <- read.csv("../data/DWC_theme_tracker.csv", header = TRUE, row.names = 1)
dim(SW_data)
```

These data could be incomplete and could benefit from outside assistance. I am slightly concerned by the lack of "rare" themes in the data set. Rare things can be important in ecology but likely won't have a big impact on the analyses I conducted. With that said, I could still use some help! Please contribute to the theme tracker. There are a few ways you could contribute:

* Edit a [Google Doc](https://docs.google.com/spreadsheets/d/1zN0OdNH8btShE3pk09FUaM7Z-UhRg9gZdgxe_IKWkfY/edit?usp=sharing)
* Submit a pull request through [`github`](https://github.com/butterflyology/Star_Wars_musical_ecology) (for those with technical skills).
* Email me through my [website](https://butterflyology.github.io/).
* Contact me on twitter: [\@butterflyology](https://twitter.com/butterflyology)


## General Plots


### All the themes

Let's make a histogram where the total number of appearances each theme makes in the saga is plotted.

Why did I make this plot?

Well, seeing how many times a theme appears in all of the *Star Wars* films can give us an idea of what the major themes are in the series (**Figure 2**). For example, *The Force / Obi Wan Kenobi's theme* is used 135 times in the films; which is 42 times more than we hear the *Main title / Luke's theme*. To really investigate the plot I made, hover your cursor over each bar to see what it represents. You can also click and drag over an area to zoom in (double-click to zoom out).


```{r plot_all, echo = FALSE, fig.cap = "**Figure 2**: Plot of all theme appearances", warning = FALSE, message = FALSE, error = FALSE}
#SW_tidy <- SW_data
SW_tidy <- as.tibble(SW_data)

SW_themes_sum <- SW_tidy %>% 
  gather(key = "Theme") %>%
  mutate(value = as.integer(value)) %>%
  group_by(Theme) %>% 
  summarize(SUM = sum(value)) %>% 
  arrange(desc(SUM))

all_themes_ggplot <- ggplot(SW_themes_sum, aes(x = reorder(Theme, desc(SUM)), y = SUM)) + 
geom_histogram (fill = SW_palette("ROTJ", n = 52, type = "continuous"), stat = "identity") + 
  theme_bw() +
  xlab("") +
  ylab("Appearances") +
  theme(axis.text.x = element_blank())

all_themes_plotly <- ggplotly(all_themes_ggplot)
plotly_all_themes_str <- plotly_build(all_themes_plotly)

#names(plotly_all_themes_str$x$data[[1]])
plotly_all_themes_str$x$data[[1]]$text <- "Qui Gon's theme - 1"
plotly_all_themes_str$x$data[[2]]$text <- "Kylo's theme #2 - 15"
plotly_all_themes_str$x$data[[3]]$text <- "Jango's Escape - 1"
plotly_all_themes_str$x$data[[4]]$text <- "March of the Resistance - 15"
plotly_all_themes_str$x$data[[5]]$text <- "Emperor's theme - 15"
plotly_all_themes_str$x$data[[6]]$text <- "The Rebels - 13"
plotly_all_themes_str$x$data[[7]]$text <- "Troopers moving - 2"
plotly_all_themes_str$x$data[[8]]$text <- "Death Star - 11"
plotly_all_themes_str$x$data[[9]]$text <- "Guardians of the Whills - 7"
plotly_all_themes_str$x$data[[10]]$text <- "Anakin's theme - 7"
plotly_all_themes_str$x$data[[11]]$text <- "Camino - 6"
plotly_all_themes_str$x$data[[12]]$text <- "Luke and Leia's theme - 7"
plotly_all_themes_str$x$data[[13]]$text <- "Battle Preparations - 4"
plotly_all_themes_str$x$data[[14]]$text <- "Duel of the Fates - 4"
plotly_all_themes_str$x$data[[15]]$text <- "The Master Switch - 4"
plotly_all_themes_str$x$data[[16]]$text <- "Jedi Steps - 4"

plotly_all_themes_str$x$data[[17]]$text <- "Rose's theme - 17"
plotly_all_themes_str$x$data[[18]]$text <- "Snoke - 8"
plotly_all_themes_str$x$data[[19]]$text <- "Rogue One Theme - 2"
plotly_all_themes_str$x$data[[20]]$text <- "Droid's theme (TESB) - 11"
plotly_all_themes_str$x$data[[21]]$text <- "Poe's theme - 8"
plotly_all_themes_str$x$data[[22]]$text <- "The Message - 4"
plotly_all_themes_str$x$data[[23]]$text <- "Falcon theme - 8"
plotly_all_themes_str$x$data[[24]]$text <- "Separatist Conspiracy theme - 11"
plotly_all_themes_str$x$data[[25]]$text <- "Battle of the Heroes - 8"
plotly_all_themes_str$x$data[[26]]$text <- "TIE Fighter attack - 4"
plotly_all_themes_str$x$data[[27]]$text <- "Arena Monsters - 8"
plotly_all_themes_str$x$data[[28]]$text <- "Jar Jar's theme - 2"
plotly_all_themes_str$x$data[[29]]$text <-  "Scarif Battle - 10"
plotly_all_themes_str$x$data[[30]]$text <- "Jedha / Saw - 9"
plotly_all_themes_str$x$data[[31]]$text <- "Vader's theme - 19"
plotly_all_themes_str$x$data[[32]]$text <- "Trade Federation - 4"
plotly_all_themes_str$x$data[[33]]$text <- "Finn's theme - 9"
plotly_all_themes_str$x$data[[34]]$text <- "Tusken Slaughter - 3"
plotly_all_themes_str$x$data[[35]]$text <- "Droid's theme (ROTJ) - 3"
plotly_all_themes_str$x$data[[36]]$text <- "Grievous' theme - 3"
plotly_all_themes_str$x$data[[37]]$text <- "Krennic's Theme - 20"
plotly_all_themes_str$x$data[[38]]$text <- "Scherzo for X-Wings - 3"
plotly_all_themes_str$x$data[[39]]$text <- "Hope - 3"
plotly_all_themes_str$x$data[[40]]$text <- "Jaba's theme - 3"
plotly_all_themes_str$x$data[[41]]$text <- "Rebel Action Theme - 3"
plotly_all_themes_str$x$data[[42]]$text <- "Kylo's theme #1 - 22"
plotly_all_themes_str$x$data[[43]]$text <- "Jyn's Theme - 23"
plotly_all_themes_str$x$data[[44]]$text <-"The Force Theme / Obi Wan Kenobi's theme - 135"
plotly_all_themes_str$x$data[[45]]$text <- "Main Title / Luke's theme- 93"
plotly_all_themes_str$x$data[[46]]$text <- "Imperial March - 79"
plotly_all_themes_str$x$data[[47]]$text <- "Yoda's theme - 25"
plotly_all_themes_str$x$data[[48]]$text <- "Rebel Fanfare - 54"
plotly_all_themes_str$x$data[[49]]$text <- "Rey's theme - 48"
plotly_all_themes_str$x$data[[50]]$text <- "Han Solo and the Princess - 33"
plotly_all_themes_str$x$data[[51]]$text <- "Across the Stars - 31"
plotly_all_themes_str$x$data[[52]]$text <- "Leia's theme - 31"
plotly_all_themes_str
```


### Themes by Film

It may also be informative to look at the distribution of themes within each film. I've made a plot where each film is represented by a bar it is filled according to the frequency of the themes in that movie. To explore this figure, hover your cursor over a bar to see the theme and number of times it appeared in that film. Try clicking on `compare data on hover` to see all the themes at once. The color for each theme is consistent across all the films. 

```{r stacked_histo, echo = FALSE, warning = FALSE, message = FALSE, fig.cap = "**Figure 3**: Themes by film"}
SW_tidy_2 <- SW_tidy %>% 
  mutate(Film = as.factor(rownames(SW_data))) %>% 
  group_by(Film) %>% 
  gather(key = Theme, value = value, -Film) %>% 
  mutate(Theme = as.factor(Theme)) %>% 
  filter(value >= 1) %>% 
  arrange(Theme)

themes_by_film <- ggplot(SW_tidy_2, aes(x = Film, y = value, fill = Theme)) + 
  geom_bar(stat = "identity", position = "stack") +
  theme_bw() +
  theme(legend.position = "none") +
  ylab("Count")

themes_film <- ggplotly(themes_by_film)
themes_film_str <- plotly_build(themes_film)

# names(themes_film_str$x$data[[1]])
# Need to do this with RegEx instead
themes_film_str$x$data[[1]]$text[1] <- "Across The Stars- 25"
themes_film_str$x$data[[1]]$text[2] <- "Across The Stars - 6"
themes_film_str$x$data[[2]]$text[1] <- "Anakin's theme - 3"
themes_film_str$x$data[[2]]$text[2] <- "Anakin's theme - 3"
themes_film_str$x$data[[2]]$text[3] <- "Anakin's theme - 1"
themes_film_str$x$data[[3]]$text[1] <- "Arena Monsters - 6"
themes_film_str$x$data[[3]]$text[2] <- "Arena Monsters - 2"
themes_film_str$x$data[[4]]$text[1] <- "Battle of the Heroes - 8"
themes_film_str$x$data[[5]]$text[1] <- "Battle Preparations - 4"
themes_film_str$x$data[[6]]$text[1] <- "Camino - 6"
themes_film_str$x$data[[7]]$text[1] <- "Death Star - 1"
themes_film_str$x$data[[7]]$text[2] <- "Death Star - 7"
themes_film_str$x$data[[7]]$text[3] <- "Death Star - 3"
themes_film_str$x$data[[8]]$text[1] <- "Droid theme (EV) - 11"
themes_film_str$x$data[[9]]$text[1] <- "Droid theme (EVI) - 3"
themes_film_str$x$data[[10]]$text[1] <- "Duel of the Fates - 1"
themes_film_str$x$data[[10]]$text[2] <- "Duel of the Fates - 1"
themes_film_str$x$data[[10]]$text[3] <- "Duel of the Fates - 3"
themes_film_str$x$data[[11]]$text[1] <- "Emperor's theme - 6"
themes_film_str$x$data[[11]]$text[2] <- "Emperor's theme - 3"
themes_film_str$x$data[[11]]$text[3] <- "Emperor's theme - 6"
themes_film_str$x$data[[11]]$text[4] <- "Emperor's theme - 3"
themes_film_str$x$data[[12]]$text[1] <- "Falcon theme - 8"
themes_film_str$x$data[[13]]$text[1] <- "Finn's theme - 9"
themes_film_str$x$data[[14]]$text[1] <- "Force theme - 11"
themes_film_str$x$data[[14]]$text[2] <- "Force theme - 8"
themes_film_str$x$data[[14]]$text[3] <- "Force theme - 20"
themes_film_str$x$data[[14]]$text[4] <- "Force theme - 18"
themes_film_str$x$data[[14]]$text[5] <- "Force theme - 14"
themes_film_str$x$data[[14]]$text[6] <- "Force theme - 19"
themes_film_str$x$data[[14]]$text[7] <- "Force theme - 13"
themes_film_str$x$data[[14]]$text[8] <- "Force theme - 25"
themes_film_str$x$data[[14]]$text[8] <- "Force theme - 7"
themes_film_str$x$data[[15]]$text[1] <- "Grievous' theme - 3"
themes_film_str$x$data[[16]]$text[1] <- "Guardians of the Whills - 7"
themes_film_str$x$data[[17]]$text[1] <- "Han Solo and the Princess - 19"
themes_film_str$x$data[[17]]$text[2] <- "Han Solo and the Princess - 9"
themes_film_str$x$data[[17]]$text[3] <- "Han Solo and the Princess - 3"
themes_film_str$x$data[[17]]$text[4] <- "Han Solo and the Princess - 2"
themes_film_str$x$data[[18]]$text[1] <- "Hope - 3"
themes_film_str$x$data[[19]]$text[1] <- "Imperial March - 1"
themes_film_str$x$data[[19]]$text[2] <- "Imperial March - 2"
themes_film_str$x$data[[19]]$text[3] <- "Imperial March - 13"
themes_film_str$x$data[[19]]$text[4] <- "Imperial March - 37"
themes_film_str$x$data[[19]]$text[5] <- "Imperial March - 20"
themes_film_str$x$data[[19]]$text[6] <- "Imperial March - 2"
themes_film_str$x$data[[19]]$text[7] <- "Imperial March - 3"
themes_film_str$x$data[[19]]$text[8] <- "Imperial March - 3"
themes_film_str$x$data[[20]]$text[1] <- "Jaba's theme - 1"
themes_film_str$x$data[[20]]$text[2] <- "Jaba's theme - 2"
themes_film_str$x$data[[21]]$text[1] <- "Jango's Escape - 1"
themes_film_str$x$data[[22]]$text[1] <- "Jar Jar's theme - 2"
themes_film_str$x$data[[23]]$text[1] <- "Jedha - Saw - 9"
themes_film_str$x$data[[24]]$text[1] <- "Jedi Steps - 1"
themes_film_str$x$data[[24]]$text[2] <- "Jedi Steps - 3"
themes_film_str$x$data[[25]]$text[1] <- "Jyn's theme - 23"
themes_film_str$x$data[[26]]$text[1] <- "Krennic's theme - 20"
themes_film_str$x$data[[27]]$text[1] <- "Kylo's theme #1 - 10"
themes_film_str$x$data[[27]]$text[2] <- "Kylo's theme #1 - 12"
themes_film_str$x$data[[28]]$text[1] <- "Kylo's theme #2 - 2"
themes_film_str$x$data[[28]]$text[2] <- "Kylo's theme #2 - 13"
themes_film_str$x$data[[29]]$text[1] <- "Leia's theme - 2"
themes_film_str$x$data[[29]]$text[2] <- "Leia's theme - 8"
themes_film_str$x$data[[29]]$text[3] <- "Leia's theme - 3"
themes_film_str$x$data[[29]]$text[4] <- "Leia's theme - 4"
themes_film_str$x$data[[29]]$text[5] <- "Leia's theme - 4"
themes_film_str$x$data[[29]]$text[6] <- "Leia's theme - 10"
themes_film_str$x$data[[30]]$text[1] <- "Luke and Leia - 1"
themes_film_str$x$data[[30]]$text[2] <- "Luke and Leia - 5"
themes_film_str$x$data[[30]]$text[3] <- "Luke and Leia - 1"
themes_film_str$x$data[[31]]$text[1] <- "Main Title - 6"
themes_film_str$x$data[[31]]$text[2] <- "Main Title - 3"
themes_film_str$x$data[[31]]$text[3] <- "Main Title - 4"
themes_film_str$x$data[[31]]$text[4] <- "Main Title - 17"
themes_film_str$x$data[[31]]$text[5] <- "Main Title - 28"
themes_film_str$x$data[[31]]$text[6] <- "Main Title - 18"
themes_film_str$x$data[[31]]$text[7] <- "Main Title - 13"
themes_film_str$x$data[[31]]$text[8] <- "Main Title - 2"
themes_film_str$x$data[[31]]$text[9] <- "Main Title - 2"
themes_film_str$x$data[[32]]$text[1] <- "March of the Resistance - 5"
themes_film_str$x$data[[32]]$text[2] <- "March of the Resistance - 10"
themes_film_str$x$data[[33]]$text[1] <- "Master Switch - 4"
themes_film_str$x$data[[34]]$text[1] <- "Poe's theme - 4"
themes_film_str$x$data[[34]]$text[2] <- "Poe's theme - 4"
themes_film_str$x$data[[35]]$text[1] <- "Qui Gon's theme - 1"
themes_film_str$x$data[[36]]$text[1] <- "Rebel Action - 3"
themes_film_str$x$data[[37]]$text[1] <- "Rebel Fanfare - 3"
themes_film_str$x$data[[37]]$text[2] <- "Rebel Fanfare - 13"
themes_film_str$x$data[[37]]$text[3] <- "Rebel Fanfare - 6"
themes_film_str$x$data[[37]]$text[4] <- "Rebel Fanfare - 9"
themes_film_str$x$data[[37]]$text[5] <- "Rebel Fanfare - 8"
themes_film_str$x$data[[37]]$text[6] <- "Rebel Fanfare - 9"
themes_film_str$x$data[[37]]$text[7] <- "Rebel Fanfare - 6"
themes_film_str$x$data[[38]]$text[1] <- "Rey's theme- 25"
themes_film_str$x$data[[38]]$text[2] <- "Rey's theme- 23"
themes_film_str$x$data[[39]]$text[1] <- "Rogue theme - 2"
themes_film_str$x$data[[40]]$text[1] <- "Rose's theme - 17"
themes_film_str$x$data[[41]]$text[1] <- "Scarif battle - 10"
themes_film_str$x$data[[42]]$text[1] <- "Scherzo for X-Wings - 3"
themes_film_str$x$data[[43]]$text[1] <- "Separatist Conspiracy - 11"
themes_film_str$x$data[[44]]$text[1] <- "Snoke - 1"
themes_film_str$x$data[[44]]$text[2] <- "Snoke - 7"
themes_film_str$x$data[[45]]$text[1] <- "The Message - 2"
themes_film_str$x$data[[45]]$text[2] <- "The Message - 2"
themes_film_str$x$data[[46]]$text[1] <- "The Rebels - 13"
themes_film_str$x$data[[47]]$text[1] <- "TIE Fighter Attack - 1"
themes_film_str$x$data[[47]]$text[2] <- "TIE Fighter Attack - 3"
themes_film_str$x$data[[48]]$text[1] <- "Trade Federation - 1"
themes_film_str$x$data[[48]]$text[2] <- "Trade Federation - 2"
themes_film_str$x$data[[48]]$text[3] <- "Trade Federation - 1"
themes_film_str$x$data[[49]]$text[1] <- "Troopers moving - 1"
themes_film_str$x$data[[49]]$text[2] <- "Troopers moving - 1"
themes_film_str$x$data[[50]]$text[1] <- "Tuskan Slaughter - 3"
themes_film_str$x$data[[51]]$text[1] <- "Vader's theme - 17"
themes_film_str$x$data[[51]]$text[2] <- "Vader's theme - 2"
themes_film_str$x$data[[52]]$text[1] <- "Yoda's theme - 1"
themes_film_str$x$data[[52]]$text[2] <- "Yoda's theme - 2"
themes_film_str$x$data[[52]]$text[3] <- "Yoda's theme - 1"
themes_film_str$x$data[[52]]$text[4] <- "Yoda's theme - 14"
themes_film_str$x$data[[52]]$text[5] <- "Yoda's theme - 4"
themes_film_str$x$data[[52]]$text[6] <- "Yoda's theme - 3"

# themes_film_str$x$data[[35]]$marker$color
for(i in 1:52){
  coloring <- SW_palette("ROTS", type = "continuous", n = 52)
  themes_film_str$x$data[[i]]$marker$color <- coloring[i]
}

themes_film_str
```

I knew it! **The Last Jedi** has the most number of thematic appearances at over 150 (**Figure 3**)! I bet if we look at the diversity metrics below the same will be true for the total number of themes present in the film.


## Analysis


### Clustering

Now we'll make a tree depicting the relationships between the eight films of the *Star Wars* saga and **Rogue One**, just as we did in the toy example above.

```{r SW_clust, include = FALSE}
SW_dist <- vegdist(SW_data, method = "jaccard", diag = FALSE, na.rm = TRUE)

SW_clust <- hclust(SW_dist, method = "complete")
```

A prediction on the clustering analysis. I postulate that the three *original trilogy* films will cluster separately from the *prequel trilogy* films (which will also cluster together). I also predict that the films of the *new trilogy* will be more similar to the *original trilogy* than the prequels.

Adding the data from **Rogue One** allows us to see where that film lies in relation to the others. Michael Giacchino rooted the music for **Rogue One** firmly within *Star Wars*. He used parts from **A New Hope** to form the themes used in **Rogue One**, for example Jyn Erso's Suite was based on "the Message," which plays in the background when Obi-Wan says "You must learn the ways of the Force...." It is also the only *Star Wars* film to share "Darth Vader's" theme with *A New Hope.*

I have also added data from two viewings of **The Last Jedi**, which I think is a masterful score. I picked up on themes for Finn and Rose, as well as the return of both of Kylo Ren's themes, *Rey's theme*, *Poe's theme*, *The March of the Resistance*, and so much more. Not going to lie, I choked up when I heard *Luke and Leia's theme* in the final scene. 


```{r SW_plot, echo = FALSE, fig.cap = "**Figure 4**: Clustering of the Star Wars films based on the their musical theme counts."}
plot(SW_clust, main = "", xlab = "", sub = "")
```

This plot (**Figure 4**) shows that the *prequel trilogy* films do indeed cluster together, and that the *original trilogy* films cluster together. When I include the data from **The Last Jedi** we see it falls out right next to **The Force Awakens**, suggesting that the *new trilogy* films have a lot in common musically. Notice that **Rogue One** is right there in between the *prequel trilogy* and the *original trilogy*? That makes a lot of sense to me because it shares some themes with the *original trilogy* but is really a uniquely scored film.


```{r DW_d, include = FALSE}
# A simple function to calculate a range of q values for a, b, and g diversity
D.iter.q <- function(data, level, q){
Spoon <- matrix(data = NA, ncol = 1, nrow = 6)
  for (i in 1:6){
  temp <- d(data, lev = level, q = i - 1, boot = FALSE)
  Spoon[i, 1] <- temp[[1]]
  }
  return(Spoon)
}

EIa <- D.iter.q(data = SW_data[1, ], level = "alpha", q = 5)
EIIa <- D.iter.q(data = SW_data[2, ], level = "alpha", q = 5)
EIIIa <- D.iter.q(data = SW_data[3, ], level = "alpha", q = 5)
EIVa <- D.iter.q(data = SW_data[4, ], level = "alpha", q = 5)
EVa <- D.iter.q(data = SW_data[5, ], level = "alpha", q = 5)
EVIa <- D.iter.q(data = SW_data[6, ], level = "alpha", q = 5)
EVIIa <- D.iter.q(data = SW_data[7, ], level = "alpha", q = 5)
EVIIIa <- D.iter.q(data = SW_data[8, ], level = "alpha", q = 5)
ROnea <- D.iter.q(data = SW_data[9, ], level = "alpha", q = 5)
```


### Jost's D

This metric is a way of counting how many things there are in a certain habitat. The cool thing about *Jost's D* is that you can consider how many things there are while accounting for how rare they are (that is the *q* on the bottom of **Figure 5**.
Here we count the number of different themes by film and consider how many different themes there are if we weight "rarity." 

```{r alpha_plot, echo = FALSE, fig.cap = "**Figure 5**: Plot of the effective number of themes by Star Wars film"}
plot(x = seq(from = 0, to = 5, length.out = 4), y = seq(from = 4, to = 20, length.out = 4), xaxt = "n", type = "n", ylab = "", xlab = "", las = 1, main = "")
axis(1, at = c(0, 1, 2, 3, 4, 5))
mtext(expression(paste(italic(alpha))), side = 2, padj = -2.5, srt = 0, cex = 2)
mtext(expression(paste(italic("q"))), 1, padj = 2.5, cex = 1.5)

# EI
points(y = EIa, x = seq(0, 5), pch = 19, cex = 1, col = SW_palette("TPM")[1])
lines(y = EIa, x = seq(0, 5), pch = 19, lwd = 2, type = "l",col = SW_palette("TPM")[1])

# E2
points(y = EIIa, x = seq(0, 5), pch = 19, cex = 1, col = SW_palette("AOTC")[2])
lines(y = EIIa, x = seq(0, 5), pch = 19, lwd = 2, type = "l",col = SW_palette("AOTC")[2])

# E3
points(y = EIIIa, x = seq(0, 5), pch = 19, cex = 1, col = SW_palette("ROTS")[6])
lines(y = EIIIa, x = seq(0, 5), pch = 19, lwd = 2, type = "l",col = SW_palette("ROTS")[6])

# E4
points(y = EIVa, x = seq(0, 5), pch = 19, cex = 1, col = SW_palette("ANH")[4])
lines(y = EIVa, x = seq(0, 5), pch = 19, lwd = 2, type = "l",col = SW_palette("ANH")[4])

# E5
points(y = EVa, x = seq(0, 5), pch = 19, cex = 1, col = SW_palette("TESB")[5])
lines(y = EVa, x = seq(0, 5), pch = 19, lwd = 2, type = "l",col = SW_palette("TESB")[5])

# E6
points(y = EVIa, x = seq(0, 5), pch = 19, cex = 1, col = SW_palette("ROTJ")[3])
lines(y = EVIa, x = seq(0, 5), pch = 19, lwd = 2, type = "l",col = SW_palette("ROTJ")[5])

# E7
points(y = EVIIa, x = seq(0, 5), pch = 19, cex = 1, col = SW_palette("TFA")[4])
lines(y = EVIIa, x = seq(0, 5), pch = 19, lwd = 2, type = "l",col = SW_palette("TFA")[4])

#E8
points(y = EVIIIa, x = seq(0, 5), pch = 19, cex = 1, col = SW_palette("TFA")[7])
lines(y = EVIIIa, x = seq(0, 5), pch = 19, lwd = 2, type = "l",col = SW_palette("TFA")[7])

#ROne
points(y = ROnea, x = seq(0, 5), pch = 19, cex = 1, col = SW_palette("Inquisitor")[1])
lines(y = ROnea, x = seq(0, 5), pch = 19, lwd = 2, type = "l",col = SW_palette("Inquisitor")[1])

legend("topright", bty = "n", legend = c("E I", "E II", "E III", "E IV", "E V", "E VI", "E VII", "E VIII", "ROne"), pch = 19, col = c(SW_palette("TPM")[1], SW_palette("AOTC")[2], SW_palette("ROTS")[6], SW_palette("ANH")[4], SW_palette("TESB")[5], SW_palette("ROTJ")[5], SW_palette("TFA")[4], SW_palette("TFA")[7], SW_palette("Inquisitor")[1]))
```

To read this plot we look at the y (vertical) axis to see the number of themes. The Greek letter alpha ($\alpha$) is the statistical designation for "unique things." Along the x (horizontal) axis we have the different weights we place on "rarity," the *q* that I mentioned above. A weight of 0 means that all themes are equal and it represents the total number of themes present in each film. As we move right along the x-axis we decrease the number of themes because we give them less weight. All the way to the right (*q* = 5) we hardly consider the effect that rare themes have on the number of themes. 

Note that **The Empire Strikes Back** actually has the fewest total number of themes (when *q* = 0) at 8, followed by **A New Hope** with 9. When we get to **The Last Jedi** there are 18 different themes that appear in the film!  **Rogue One** actually have the highest total number of themes, but when we care less about rare themes (*q* = 5) **The Last Jedi** is a bit higher. One thing that appears evident from this analysis, is that all *Star Wars* films have ~5 themes that we hear frequently in each film. 

One last note of geekery. The colors from that plot were made with an `R` package called [`spaceMovie`](https://github.com/butterflyology/spaceMovie), which uses colors from the *Star Wars* franchise.


### NMDS

Lastly, I employ another method of visualization called **NMDS** (**N**on-**M**etric Multi**D**imensional **S**caling) which  plots the locations of each "habitat" in ordination space. In this case, each film appears on the plot in a place relative to the other films. That is to say, similar things should be closer together than dissimilar things. 

```{r nmds, include = FALSE}
series <- c("Prequel", "Prequel", "Prequel", "Original Trilolgy", "Original Trilolgy", "Original Trilolgy", "New Trilogy", "New Trilogy", "Story")
SW_nmds <- metaMDS(SW_dist, trace = FALSE)
stressplot(SW_nmds, SW_dist)
```

```{r nmds_plot, echo = FALSE, results = "hide", warning = FALSE, error = FALSE, message = FALSE, fig.cap = "**Figure 6**: NMDS Ordination plot of the Star Wars films."}
plot(SW_nmds, type = "t", las = 1, cex = 1.2, xlab = "Axis 1", ylab = "Axis 2", xlim = c(-0.5, 0.5), ylim = c(-0.55, 0.4))
ordiellipse(SW_nmds, series, draw = "lines", col = c(alpha(SW_palette("TPM")[2], 0.3), c(alpha(SW_palette("ANH")[4], 0.3)),
c(alpha(SW_palette("TESB")[1], 0.3))), kind = "ehull", lwd = 3)
```

Think about which films you could draw an ellipse around without including any other films (**Figure 6**). We could have the computer draw an ellipse around the *prequel trilogy* so that it only contains the prequels,  suggesting that these films are more similar to each other than they are to other films. We can also have the computer draw an ellipse around the *original trilogy*. Lastly, **The Force Awakens** and **The Last Jedi** are off by themselves and I predict that once **Episode IX** comes out it fall out with them. Right now the computer can't draw an ellipse around two points. As in the clustering analysis, **Rogue One** is off doing its own thing. All together, these findings are consistent with the clustering "tree" we saw earlier. 


## Conclusions

I have four big takeaways about the music of the *Star Wars* films based on this exercise:

1. The *original trilogy* films are most similar to each other.
1. The *prequel trilogy* films are most similar to each other.
1. **The Force Awakens** and **The Last Jedi** are grouping together but appear close to the *original trilogy*.
1. **Rogue One** is its own thing, but more similar to the *original trilogy* than anything else.

These results make a lot of sense to me. I interpret this to mean that John Williams kept similar themes throughout each of the two trilogies, and that the *new trilogy* is building off of the *original trilogy*. I also see that Michael Giacchino used themes found in **A New Hope** to ground **Rogue One** in the *Star Wars* musical universe, but made the score his own.

Before the **The Last Jedi** premiered, I predicted that **Episode VIII** would be closely related to **The Force Awakens**. I'm glad to see that I was right. If you don't believe that I predicted this, go through the "history" in the code repository that houses this page and see for yourself. Lastly, 


## Table

In case you didn't want to follow the links to find the data I used for this post, below is copy:

```{r kable, echo = FALSE}
knitr::kable(x = SW_data, align = "c", format = "html", caption = "**Table 2**: Data used in the present analysis") %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "responsive")) %>% 
  scroll_box(width = "750px", height = "450px")
```